Counterfactual Training

### 2024-12-18 paper/presenations/slides.qmd
uff

$$
\ell_{\text{clf}}(f(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \text{r}
$$

- $x^\prime$ are counterfactuals of $x$.
- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
- Counterfactuals are recycled as adversarial examples.

@@

1. Generate `nce` counterfactuals and distribute across mini-batches.
2. For each batch compute:
    - Classifier loss: $\ell_{\text{clf}}(f(x),y)$
    - Generator loss: $\ell_{\text{gen}}(x^\prime,x)$
    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y)$
3. Backpropagate all losses and update parameters.

>>>>

$$
\ell_{\text{clf}}(f_\theta(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x;\theta) + \lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)
$$

- $x^\prime$ are counterfactuals of $x_{s}\subseteq x$.
- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
- Counterfactuals are recycled as adversarial examples.

@@

1. Generate `nce` counterfactuals and distribute across mini-batches.
2. For each batch compute:
    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
    - Generator loss: $\ell_{\text{gen}}(x^\prime,x;\theta)$
    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)$
3. Backpropagate all losses and update parameters.

## Motivation and Intuition

- Instead of using SGLD to sample from $p(x|t;\theta)$, we use counterfactual generators.
- The idea is to align counterfactual explanations with observed data to induce plausibility.
- This should only work if counterfactuals are generated faithfully (favorable evidence).
- Approach can be leveraged to impicitly encode mutability and domain constraints in model.
